\LARGE \bf Nonparametric Distribution Models for Predicting and \\ Managing Computational Performance Variability*

### 2018-02-13 Paper_[2018-01-20]_IEEESE/paper.tex
Finalized some more modifications on the ACM paper. Ready to deliver to Dr. Watson for further review.

\author{Thomas C. H. Lux$^{1}$, Layne Watson$^{2}$, Jon Bernard$^{1}$, Tyler H. Chang$^{1}$, Bo Li$^{1}$, \\ Li Xu$^{3}$, Godmar Back$^{2}$, Ali R. Butt$^{2}$, Kirk W. Cameron$^{2}$, Yili Hong$^{4}$% <-this % stops a space
\thanks{*This work was supported by the National Science Foundation}%
\thanks{$^{1}$T. Lux, J. Bernard, T. Chang, B. Li are doctoral students of Computer Science at Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
\thanks{$^{2}$L. Watson, G. Back, A. Butt, and K. Cameron are Faculty of Computer Science at Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
\thanks{$^{3}$L. Xu is a doctoral student of Statistics at Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
\thanks{$^{4}$Y. Hong is with the Faculty of Statistics at Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Performance variability can have a significant impact on many applications of computing. Cloud Computing, High Performance Computing, and Computer Security communities each exert considerable effort managing and analyzing variability throughout the system stack. This work presents and evaluates a methodology for predicting precise characteristics of the computational performance variability of an input/output (I/O) application over varying system configurations. Results demonstrate that the presented methodology is capable of precisely modeling performance variability, which could allow applications that tighten service level agreements, maximize computational throughput, and obfuscate system configurations against malicious users.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Predicting performance variability in a computer system is a challenging problem that has primarily been attempted in one of two ways: (1) build a statistical model of the performance data collected by running experiments on the system at select settings, or (2) run artificial experiments using a simplified simulation of the target system to estimate architecture and application bottlenecks. In this paper, the proposed modeling techniques rest in the first category and they represent a notable increase in the ability to model precise characteristics of variability.

@@

Performance and its variability can be summarized by a variety of statistics. Mean, range, standard deviation, variance, and interquartile range are a few summary statistics that describe performance and variability. However, the most precise characterization of any ordinal performance metric is the cumulative distribution function (CDF), or its derivative the probability density function (PDF). Previous techniques for predicting system performance have strictly modeled real-valued summary statistics because there exists a large base of mathematical techniques capable of approximating functions of the form $f: \mathbb{R}^d \rightarrow \mathbb{R}$. There is a smaller theoretical foundation for approximating functions of the form $f: \mathbb{R}^d \rightarrow g$ with $g: \mathbb{R} \rightarrow \mathbb{R}$. The latter is the approximation of functions.

A major hurdle when modeling functions such as the CDF or PDF, is that certain properties must be maintained in order for the functions to remain valid. It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) > 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F'(x) \geq 0$. For the purpose of function approximation, the properties of the CDF are more straightforward to maintain. This work utilizes the fact that a convex combination of CDFs results in a valid CDF. Given $G(x) = \sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid CDF, $G$ must also be a valid CDF. A demonstration of how this is applied can be seen in Figure \ref{fig:prediction_example}. Next, the three techniques used to model CDFs are summarized, a mechanism for identifying important features is proposed, and the error measure used throughout this work is presented. For each of the following, $n \in \mathcal{N}$ refers to the number of training samples (known system configurations), $d \in \mathcal{N}$ refers to the dimension of the data (ordinal system parameters), $x^{(i)} \in \mathbb{R}^d$ is a known system configuration with known CDF $F_{x^{(i)}}$, and $y \in \mathbb{R}^d$ refers to a previously unseen system configuration with CDF $F_y$.

@@

The Delaunay method of interpolation is a well studied geometric technique for producing an interpolant \cite{lee1980two}. The Delaunay triangulation of a set of data points into simplices is such that the sphere defined by the vertices of each simplex contains no data points in the sphere's interior. For a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$, $x^{(d)}$, and functions $F_{x^{(i)}}$, $i=0$, $\ldots$, $d$, $y$ is a unique convex combination of the vertices,

@@

and the Delaunay interpolant $F_y$ at $x$ is given by

@@

\begin{figure}[htb]
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-delaunay-example-prediction.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this example, the general methodology for predicting a CDF and evaluating error can be seen. The Delaunay method chose three source system configurations (dotted lines) and assigned weights \{.3, .4, .3\} (top to bottom). The weighted sum of the three known CDFs produces the predicted CDF (dashed line). The KS Statistic (red arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example. This prediction is deemed a failure by p-values 0.05 and 0.01, however it is not a failure according to p-values 0.001 and 1.0e-6.
  \vspace{-.1cm}}
  \label{fig:prediction_example}
\end{figure}

The computational complexity of interpolating via Delaunay (for the implementation used) is approximately $\mathcal{O}(n d^4 log d)$, which is capable of scaling reasonably to $d \leq 50$. In the present application, Delaunay is used to select up to $d+1$ system configurations and assign weights to each in order to generate a prediction for a new system configuration.

@@

The Max Box mesh is an interpolation technique that utilizes overlapping box splines \cite{de2013box} as basis functions that are shifted and scaled to support box shaped regions. The boxes are constructed such that each box has strictly one point on its interior while maintaining the largest minimum distance between the interior point and the closest face of the box (a measure of centrality). Given a set of box splines $b^{x^{(i)}}$ with the max box properties anchored at each $x^{(i)}$,

$$ F_y = \frac{\sum_i b^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i b^{x^{(i)}}(y)}, $$

where $b^{x^{(i)}}(y) \geq 0$ and the denominator causes the sum of weights to be one. The computational complexity of interpolation via the Max Box mesh is $\mathcal{O}(n^2 d)$. The Max Box mesh interpolates a function $F_y$ by producing a convex combination of known functions $F_{x^{(i)}}$, where the weights are determined by the box splines.

@@

A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}. Nearest neighbor inherently utilizes the convex region surrounding a central control point in whose interior all points are closer to that center than any other control point, which is also called the Voronoi cell \cite{dirichlet1850reduction}. The Voronoi mesh smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$ with

$$ v^{x^{(i)}}(y) = \left(1 - \frac{||y - x^{(i)}||_2}{2 \ d(y - x^{(i)} \mid x^{(i)})} \right), $$

where $d(\ \cdot \mid x^{(i)})$ is the distance between $x^{(i)}$ and the nearest boundary of the Voronoi cell along the vector $\cdot\ $. While all basis functions $v^{x^{(i)}}(x^{(i)}) = 1$, the $2$ in the denominator causes all basis functions to go linearly to $0$ at the boundary of the twice-expanded Voronoi cell. Now we have

$$ F_y = \frac{\sum_i v^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i v^{x^{(i)}}(y)}, $$

where $v^{x^{(i)}}(y) \geq 0$ by definition and the denominator causes the sum of weights to be one. The computational complexity of interpolation via Voronoi mesh is $\mathcal{O}(n^2 d)$. The Voronoi mesh interpolates a function $F_y$ just as the Max Box mesh and Delaunay triangulation, by a convex combination of known functions $F_{x^{(i)}}$.
%% ===================================================================
\subsection{Feature Weighting}
\label{sec:feature_weighting}

Machine learning has demonstrated that an important procedure in any application of predictive methodologies is identifying those features of training data that are most relevant to making accurate predictions \cite{guyon2003introduction}. Selection strategies such as the floating searches studied in \cite{pudil1994floating} or others compared in \cite{ferri1994comparative} can be too expensive for large approximation problems. Rather, this work poses feature selection as a continuous optimization problem. Let $X$ be an $n \times d$ matrix of $n$ known system configurations with $d$ parameters each normalized to be in the unit cube. Define an error function $err: \mathbb{R}^d \rightarrow \mathbb{R}_+$ that estimates the error of a predictive model trained on $X diag(w)$, $w \in \mathbb{R}^d$,  by performing 10-fold cross validation using 80\% of the rows of $X diag(w)$ as training and 20\% as testing. A minimal solution to this error function could be considered an optimal weighting of the features of $X$. Minimization is performed using the modified simulated annealing technique presented in \cite{lux2016convergence} because of its empirically fast convergence in the absence of a readily computable gradient.

@@

The two-sample KS test is a useful nonparametric test for declaring when two CDFs do not originate from the same underlying distribution while only assuming stationarity, finite mean, and finite variance. The null hypothesis (that two CDFs come from the same underlying distribution) is rejected at level $p \in [0,1]$ when

$$ KS > \sqrt{-\frac{1}{2}ln\big(\frac{p}{2}\big)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$

with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test presented in this work $n_1 = n_2 = 150$. An example of the round-trip prediction methodology from known and predicted distributions to the calculation of error can be seen in Figure \ref{fig:prediction_example}.

@@

\begin{table}
  \centering
  \begin{tabular}{c|c}
    \hline
    \textbf{System Parameter} & \textbf{Values}\\
    \hline
    File Size (KB) & 4, 16, 64, 256, 1024, 4096, 8192, 16384\\
    \hline
    Record Size (KB) & \multilinecell{4, 8, 16, 32, 64, 128, 256, 512,\\ 1024, 2048, 4096, 8192, 16384}\\
    \hline
    Thread Count & 1, 8, 16, 24, 32, 40, 48, 56, 64\\
    \hline
    Frequency (GHz) & 1.2, 1.6, 2, 2.3, 2.8, 3.2, 3.5\\
    \hline
    Test Type & \multilinecell{Readers, Rereaders, Random Readers, \\ Initial Writers, Rewriters, Random Writers}\\
    \hline
  \end{tabular}
  \caption{A description of system parameters considered for IOzone. Record size must be less than file size during execution.
    \vspace{-.5cm}}
  \label{tab:data_description}
\end{table}

@@

This paper presents a variability modeling case study with a five-dimensional dataset produced by executing the IOzone benchmark \cite{iozone} on a homogeneous cluster of computers. Each node contains two Intel Xeon E5-2637 CPUs offering a total of 16 CPU cores with 16GB of DRAM. While the CPU frequency varies depending on the test configuration, the I/O from IOzone is performed by an ext4 filesystem sitting above an Intel SSDSC2BA20 SSD drive. At the time of data collection, a Linux kernel version of 4.13.0 was used. The system performance data was collected over two weeks by executing IOzone 150 times for each of a select set of approximately 18K system configurations, for a total of approximately 2.7M executions of IOzone. A single IOzone execution reports the max I/O throughput in kilobytes per second seen for the selected test. The summary of the data generated in the experiments for this paper can be seen in Table \ref{tab:data_description}. Distributions of raw throughput values being modeled can be seen in Figure \ref{fig:throughput_histogram}.

Some mild preprocessing was necessary to prepare the data for modeling and analysis. All features were shifted by their minimum value and scaled by their range, mapping each feature independently into $[0,1]$. This normalization ensures each feature is treated equally by the interpolation techniques and should be performed on all data before building models and making predictions regardless of application. All 150 repeated trials for a system configuration were grouped with that configuration. The only non-ordinal feature in this data is the test. All tests were treated as different applications and were separated for modeling and analysis. (i.e. predictions for the ``readers'' test were made using known configurations for the ``readers'' test only.) The division of non-ordinal features into separate models is necessary for the presented methodology, however future improvements are discussed in Section \ref{sec:discussion}.

@@

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.5\textwidth,trim={1.5cm 0 0 0}]{plot-histogram-80_20-KS.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histograms of the prediction error averaged over 10 folds for each modeling algorithm when trained with 80\% of the data. The distributions show the KS statistics for the predicted throughput distribution versus the actual throughput distribution. The four vertical red lines represent commonly used p-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively. All predictions to the right of a red line represent failed predictions that are significantly different (by respective p-value) from the actual distribution according to the KS test.
  \vspace{-.1cm}}
  \label{fig:ks_histogram_80_20}
\end{figure}

@@

All three interpolation techniques are used to predict the distribution of I/O throughput values at previously unseen system configurations. In order to improve robustness of the error analysis, 10-fold cross validation is used. 10 random selections of 80\% of the IOzone data are used to train each model and the remaining 20\% provide estimations of approximation error for each model. The recorded errors are averaged over the 10 random folds and grouped by unique system configuration. The folds are identically seeded for each interpolation technique, ensuring consistency in the training and testing sets.

The aggregation of errors across all IOzone tests given 80\% of the data as training can be seen in Figure \ref{fig:ks_histogram_80_20}. Agglomerate errors for each technique present similarly, each resembles a Gamma distribution. The percentages of failed predictions with varying p-values are on display in Table \ref{tab:p_value_failure_rate}. The primary p-value used for analyses in this work is 0.001. This p-value is chosen because close to 2K predictions are made for each test. Also, applications executed in cloud and HPC systems that could benefit from statistical modeling will be executed at least on the order of 1000's of times. In line with this knowledge, it is important to ensure that only a small fraction of interpretable results could occur solely under the influence of random chance. When considering the p=0.001 results for each technique, a little under half of the generated predictions fail. In a traditional application of machine learning techniques a failure rate of 45\% would be a poor result, however in this situation the complexity of the problem warrants a slightly different interpretation. These predictions are of a very \textit{precise} characterization of performance variability, in fact the strongest possible characterization of variability. Globally, only a little under half of the predictions fail to capture \textit{all} of the characteristics of performance variability at new system configurations. It is also demonstrated later in this Section that this result can likely be improved.

\begin{table}
  \centering
  \begin{tabular}{c|c|c}
    \hline
    \textbf{Algorithm} & \textbf{P-Value} & \textbf{\% Failed Predictions} \\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .05 & \multilinecell{58.4\\69.3\\61.9}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .01 & \multilinecell{51.1\\58.4\\53.4}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .001 & \multilinecell{44.1\\46.9\\45.1}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & 1.0e-6 & \multilinecell{31.4\\26.6\\28.7}\\
    \hline
  \end{tabular}
  \caption{The prediction failure rate by the KS-test when provided different selections of p-values. These accompany the results from Figure \ref{fig:ks_histogram_80_20}. 
    \vspace{-.5cm}}
  \label{tab:p_value_failure_rate}
\end{table}

While interpreting failure rates of these interpolation techniques, it is important to consider at what rate the error reduces with increasing amounts of training data. Figure \ref{fig:ks_failure_by_training} displays the change in p=0.001 failure rate with increasing density of training data up to the maximum density allowed by this training set. Delaunay provides the best results with the least training data by about 5\%, but these low-density prediction error rates are unacceptably high (90\% failure). Increased density demonstrates a roughly $(1 - \sqrt{n})$ rate of reduction in error that will level off around 20-30\% with a training set twice the size of the one collected for this study. However, constructing a training set of such a size would likely require months of compute time and be infeasible for many applications. The rate of convergence seen also suggests that these models are successfully converging on the lowest possible error given the (presumed) Lipschitz constant of the data and any latent variables that may perfectly describe system I/O throughput variability.

>>>>

\author{Thomas C. H. Lux$^{1}$, Layne T. Watson$^{2}$, Jon Bernard$^{1}$, Tyler H. Chang$^{1}$, Bo Li$^{1}$, Xiaodong Yu$^{1}$,\\ Li Xu$^{3}$, Godmar Back$^{2}$, Ali R. Butt$^{2}$, Kirk W. Cameron$^{2}$, Yili Hong$^{4}$, Danfeng Yao$^{2}$% <-this % stops a space
\thanks{*This work was supported by the National Science Foundation Grant CNS-1565314.}%
\thanks{$^{1}$Doctoral student, Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA {\tt\small (tchlux at vt.edu)}}%
\thanks{$^{2}$Department of Computer Science, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
\thanks{$^{3}$Doctoral student, Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
\thanks{$^{4}$Department of Statistics, Virginia Polytechnic Institute and State University, Blacksburg, Virginia 24060, USA}%
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Performance variability can have a significant impact on many applications of computing. Cloud computing, high performance computing, and computer security communities each exert considerable effort managing and analyzing variability throughout the system stack. This work presents and evaluates a methodology for predicting precise characteristics of the computational performance variability of an input/output (I/O) application over varying system configurations. Results demonstrate that the presented methodology is capable of precisely modeling performance variability, which could allow applications that tighten service level agreements, maximize computational throughput, and obfuscate system configurations against malicious users.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Predicting performance variability in a computer system is a challenging problem that has primarily been attempted in one of two ways: (1) build a statistical model of the performance data collected by running experiments on the system at select settings, or (2) run artificial experiments using a simplified simulation of the target system to estimate architecture and application bottlenecks. In this paper, the proposed modeling techniques rest in the first category and represent a notable increase in the ability to model precise characteristics of variability.

@@

Performance and its variability can be summarized by a variety of statistics. Mean, range, standard deviation, variance, and interquartile range are a few summary statistics that describe performance and variability. However, the most precise characterization of any ordinal performance metric is the cumulative distribution function (CDF), or its derivative the probability density function (PDF). Previous techniques for predicting system performance have strictly modeled real-valued summary statistics because there exists a large base of mathematical techniques capable of approximating functions of the form $f: \mathbb{R}^d \rightarrow \mathbb{R}$. There is little systems work in approximating functions of the form $f: \mathbb{R}^d \rightarrow \{g \mid g: \mathbb{R} \rightarrow \mathbb{R}\}$.

A major hurdle when modeling functions such as the CDF or PDF is that certain properties must be maintained. It is necessary that a PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x) \geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF $F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are $F(x) \in [0,1]$ and $F(x)$ is monotone increasing. For the purpose of function approximation, the properties of the CDF are more straightforward to maintain. This work utilizes the fact that a convex combination of CDFs results in a valid CDF. Given $G(x) = \sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$ is a valid CDF, $G$ must also be a valid CDF. A demonstration of how this is applied can be seen in Figure \ref{fig:prediction_example}. Next, the three techniques used to model CDFs are summarized, a mechanism for identifying important features is proposed, and the error measure used throughout this work is presented. For each of the following, $n \in \mathcal{N}$ refers to the number of training samples (known system configurations), $d \in \mathcal{N}$ refers to the dimension of the data (ordinal system parameters), $x^{(i)} \in \mathbb{R}^d$ is a known system configuration with known performance CDF $F_{x^{(i)}}$, and $y \in \mathbb{R}^d$ refers to a previously unseen system configuration with performance CDF $F_y$.

@@

The Delaunay triangulation is a well-studied geometric technique for producing an interpolant \cite{lee1980two}. The Delaunay triangulation of a set of data points into simplices is such that there are no data points inside the sphere defined by the vertices of each simplex. For a $d$-simplex S with vertices $x^{(0)}$, $x^{(1)}$, $\ldots$, $x^{(d)}$, and functions $F_{x^{(i)}}$, $i=0$, $\ldots$, $d$, $y \in S$ is a unique convex combination of the vertices,

@@

and the Delaunay interpolant $F_y$ at $y$ is given by

@@

\begin{figure}[htb]
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-delaunay-example-prediction.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this example, the general methodology for predicting a CDF and evaluating error can be seen. The Delaunay method chose three source system configurations (dotted lines) and assigned weights \{.3, .4, .3\} (top to bottom). The weighted sum of the three known CDFs produces the predicted CDF (dashed line). The KS Statistic (red arrow) computed between the true CDF (solid line) and predicted CDF (dashed line) is 0.2 for this example. The KS test null hypothesis is rejected by $p$-value 0.01, however it is not rejected by $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction_example}
\end{figure}

The computational complexity of Delaunay interpolation (for the implementation used) is $\mathcal{O}(n d^4 \log d)$, which is capable of scaling reasonably to $d \leq 50$. In the present application, a Delaunay simplex $S$ containing $y$ is found, then the $d+1$ vertices (system configurations) of $S$ are used to assign weights to each vertex and produce the predicted CDF for system configuration $y$.

@@

The Max Box mesh is an interpolation technique that utilizes overlapping box splines \cite{de2013box} as basis functions that are shifted and scaled to have support over box shaped regions. The boxes are constructed such that each box has exactly one point in its interior while maintaining the largest minimum distance between the interior point and the closest face of the box (a measure of centrality). Given a set of box splines $\bigl\{b^{x^{(i)}}\bigr\}$ with the max box properties anchored at interior points $\bigl\{x^{(i)}\bigr\}$,

$$ F_y = \frac{\sum_i b^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i b^{x^{(i)}}(y)}. $$
Note that the box splines always satisfy $b^{x^{(i)}}(y) \geq 0$. The computational complexity of interpolation via the Max Box mesh is $\mathcal{O}(n^2 d)$.

@@

A well-studied technique for classification and approximation is the nearest neighbor algorithm \cite{cover1967nearest}. Nearest neighbor inherently utilizes the convex region $C^{x^{(i)}}$ (Voronoi cell \cite{dirichlet1850reduction}) consisting of all points closer to $x^{(i)}$ than any other point $x^{(j)}$. The Voronoi mesh smooths the nearest neighbor approximation by utilizing the Voronoi cells to define support via a generic basis function $v: \mathbb{R}^d \rightarrow \mathbb{R}_+$ given by

$$ v^{x^{(i)}}(y) = \left(1 - \frac{||y - x^{(i)}||_2}{2 \ d(y - x^{(i)} \mid x^{(i)})} \right)_+, $$
where $d(w \mid x^{(i)})$ is the distance between $x^{(i)}$ and the boundary of the Voronoi cell $C^{x^{(i)}}$ in the direction $w$. $v^{x^{(i)}}\bigl(x^{(j)}\bigr) = 0$ and $v^{x^{(i)}}$ has local support, giving the interpolated value

$$ F_y = \frac{\sum_i v^{x^{(i)}}(y) F_{x^{(i)}}}{\sum_i v^{x^{(i)}}(y)}, $$
where $0 \leq v^{x^{(i)}}(y) \leq 1$. The computational complexity of interpolation via the Voronoi mesh is $\mathcal{O}(n^2 d)$. All of the approximations are an interpolant involving a convex combination of known functions $F_{x^{(i)}}$.
%% ===================================================================
\subsection{Feature Weighting}
\label{sec:feature_weighting}

It is well-known that an important procedure in any application of predictive methodologies is identifying those features of the data that are most relevant to making accurate predictions \cite{guyon2003introduction}. Selection strategies such as the floating searches studied in \cite{pudil1994floating} or others compared in \cite{ferri1994comparative} can be too expensive for large approximation problems. Rather, this work poses feature selection as a continuous optimization problem. Let $X$ be an $n \times d$ matrix of $n$ known system configurations with $d$ parameters each normalized to be in $[0,1]$. Define an error function that computes the error of a predictive model trained on $X\hbox{ diag }w$, $w \in \mathbb{R}^d$,  by performing ten random splits with 80\% of the rows of $X\hbox{ diag }w$ for training and 20\% for testing. A minimum of this error function could be considered an optimal weighting of the features of $X$. Minimization is performed using a zero order method in the absence of a readily computable gradient.

@@

The two-sample KS test is a useful nonparametric test for comparing two CDFs while only assuming stationarity, finite mean, and finite variance. The null hypothesis (that two CDFs come from the same underlying distribution) is rejected at level $p \in [0,1]$ when

$$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test presented in this work $n_1 = n_2 = 150$. An example of the round-trip prediction methodology from known and predicted distributions to the calculation of error can be seen in Figure \ref{fig:prediction_example}.

@@

\begin{table}
  \centering
  \begin{tabular}{c|c}
    \hline
    \textbf{System Parameter} & \textbf{Values}\\
    \hline
    File Size (KB) & 4, 16, 64, 256, 1024, 4096, 8192, 16384\\
    \hline
    Record Size (KB) & \multilinecell{4, 8, 16, 32, 64, 128, 256, 512,\\ 1024, 2048, 4096, 8192, 16384}\\
    \hline
    Thread Count & 1, 8, 16, 24, 32, 40, 48, 56, 64\\
    \hline
    Frequency (GHz) & 1.2, 1.6, 2, 2.3, 2.8, 3.2, 3.5\\
    \hline
    Test Type & \multilinecell{Readers, Rereaders, Random Readers, \\ Initial Writers, Rewriters, Random Writers}\\
    \hline
  \end{tabular}
  \caption{A description of system parameters considered for IOzone. Record size must be $\leq$ file size during execution.
    \vspace{-.5cm}}
  \label{tab:data_description}
\end{table}

@@

This paper presents a variability modeling case study with a five-dimensional dataset produced by executing the IOzone benchmark \cite{iozone} on a homogeneous cluster of computers. Each node contains two Intel Xeon E5-2637 CPUs offering a total of 16 CPU cores with 16GB of DRAM. While the CPU frequency varies depending on the test configuration, the I/O from IOzone is performed by an ext4 filesystem sitting above an Intel SSDSC2BA20 SSD drive. At the time of data collection, Linux kernel Version 4.13.0 was used. The system performance data was collected over two weeks by executing IOzone 150 times for each of a select set of approximately 18K system configurations, for a total of approximately 2.7M executions of IOzone. A single IOzone execution reports the max I/O throughput in kilobytes per second seen for the selected test type. The summary of the data components in $x^{(i)}$ for the experiments for this paper can be seen in Table \ref{tab:data_description}. Distributions of raw throughput values being modeled can be seen in Figure \ref{fig:throughput_histogram}.

Some mild preprocessing was necessary to prepare the data for modeling and analysis. All features were shifted by their minimum value and scaled by their range, mapping each feature independently into $[0,1]$. This normalization ensures each feature is treated equally by the interpolation techniques and should be performed on all data before building models and making predictions regardless of application. All 150 repeated trials for a system configuration were grouped with that configuration. The only nonordinal feature in this data is the test type. All test types were treated as different applications and were separated for modeling and analysis, i.e., predictions for the ``readers'' test type were made using only known configurations for the ``readers'' test type.

@@

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.5\textwidth,trim={1.5cm 0 0 0}]{plot-histogram-80_20-KS.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {KS Statistic for Predicted vs. Actual};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {Count};
  \end{tikzpicture}
  \caption{Histograms of the prediction error for each modeling algorithm from ten random splits when trained with 80\% of the data. The distributions show the KS statistics for the predicted throughput distribution versus the actual throughput distribution. The four vertical red lines represent commonly used $p$-values \{0.05, 0.01, 0.001, 1.0e-6\} respectively. All predictions to the right of a red line represent CDF predictions that are significantly different (by respective $p$-value) from the actual distribution according to the KS test.
  \vspace{-.1cm}}
  \label{fig:ks_histogram_80_20}
\end{figure}

@@

All three interpolation techniques are used to predict the distribution of I/O throughput values at previously unseen system configurations. In order to improve robustness of the error analysis, ten random selections of 80\% of the IOzone data are used to train each model and the remaining 20\% provide approximation error for each model. The recorded errors are grouped by unique system configuration and then averaged within each group. The samples are identical for each interpolation technique, ensuring consistency in the training and testing sets.

The aggregation of errors across all IOzone tests given 80\% of the data as training can be seen in Figure \ref{fig:ks_histogram_80_20}. Agglomerate errors for each technique resemble a Gamma distribution. The percentages of significant prediction errors with varying $p$-values are on display in Table \ref{tab:p_value_failure_rate}. The primary $p$-value used for analyses in this work is 0.001, chosen because close to 2K predictions are made for each test type. Also, applications executed in cloud and HPC systems that could benefit from statistical modeling will be executed at least thousands of times. In line with this knowledge, it is important to ensure that only a small fraction of interpretable results could occur solely under the influence of random chance. When considering the $p=0.001$ results for each technique, a little under half of the predicted CDFs are significantly different from the measured (and presumed) correct CDFs. A rejection rate of 45\% would seem a poor result, however in this situation the complexity of the problem warrants a slightly different interpretation. These predictions are a very \textit{precise} characterization of performance variability, in fact the strongest possible characterization of variability. Globally, only a little under half of the predictions fail to capture \textit{all} of the characteristics of performance variability at new system configurations. It is also demonstrated later in this Section that this result can likely be improved.

\begin{table}
  \centering
  \begin{tabular}{c|c|c}
    \hline
    \textbf{Algorithm} & \textbf{$P$-Value} & \textbf{\% N.H. Rejections} \\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .05 & \multilinecell{58.4\\69.3\\61.9}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .01 & \multilinecell{51.1\\58.4\\53.4}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & .001 & \multilinecell{44.1\\46.9\\45.1}\\
    \hline
    \multilinecell{Delaunay\\Max Box Mesh\\Voronoi Mesh} & 1.0e-6 & \multilinecell{31.4\\26.6\\28.7}\\
    \hline
  \end{tabular}
  \caption{Percent of null hypothesis rejections rate by the KS-test when provided different selections of $p$-values. These accompany the percent of null hypothesis rejection results from Figure \ref{fig:ks_histogram_80_20}. 
    \vspace{-.5cm}}
  \label{tab:p_value_failure_rate}
\end{table}

While interpreting null hypothesis rejection rates for these interpolation techniques, it is important to consider how the rejection rate reduces with increasing amounts of training data. Figure \ref{fig:ks_failure_by_training} displays the change in $p=0.001$ null hypothesis rejection rate with increasing density of training data up to the maximum density allowed by this set. Delaunay interpolation provides the best results with the least training data by about 5\%, but these low density rejection rates are unacceptably high (90\%). Figure \ref{fig:ks_failure_by_training} clearly shows that this data set and/or the system variables used in the models of performance variability is inadequate to capture the full variability map from system parameters to performance CDF. Which or both obtains is not clear. A few well chosen data points can significantly improve the interpolants, and thus a careful study of the rejection instances is warranted, besides enlarging the set of system variables being modeled.

It may be misleading to consider the global performance of each prediction technique across all test types, as some test types are more difficult than others to predict and have more apparent latent variables. In Figure \ref{fig:ks_failure_by_training_and_test}, the relative difficulty of each IOzone test type can be compared. The I/O test types analyzing reads are typically approximated with lower error than those test types analyzing writes. Regardless of test type, in the aggregate results the KS statistics hover consistently around 0.15, demonstrating an impressively low KS statistic for predictions. In order to address the opacity of aggregate analysis, another case study and an application of the methodology from Section \ref{sec:feature_weighting} is presented in Table \ref{tab:optimized_p_value_failure_rate}.

\begin{figure}
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.4\textwidth,]{plot-KS-Failure-by-Training.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {Percentage of Training Data};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {\% Null Hypothesis Rejections};
  \end{tikzpicture}
  \caption{The performance of each algorithm on the KS test ($p=0.001$) with increasing amounts of training data averaged over all IOzone test types and ten random splits of the data. The training percentages range from 5\% to 95\% in increments of 5\%. Delaunay is the best performer until 95\% of data is used for training, at which Max Box mesh becomes the best performer by a fraction of a percent.
  \vspace{-.1cm}}
  \label{fig:ks_failure_by_training}
\end{figure}
